---
title: "The Brain & Mind Centre"
author: "Sydney Informatics Hub"
subtitle: Day 2
output:
  html_document:
    hightlight: github
    theme: lumen
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)
```

```{r installation, message=FALSE}
#install.packages("ggplot2")
library(ggplot2)
library(scales)
```

## Introduction to ggplot2

* ggplot2 is a data visualization package for the statistical programming language R. 
* It was created by Hadley Wickham in 2005 (while he was a graduate student at Iowa state). 
* It is based on Grammar of Graphics (Leland Wilkinson 2005), is composed of a set of independent components that can be composed in many different ways.
* Other data visualization packages in R include base and lattice. 

### What Is The Grammar Of Graphics?

Grammar of English language has components like verbs, nouns, adjectives, articles, etc to form a sentence. 

Grammar Of Graphics independently specifies plot building blocks and combine them to create just about any kind of graphical display you want. Building blocks of a graph include:

* data
* aesthetic mapping
* geometric object
* statistical transformations
* scales
* coordinate system
* position adjustments
* faceting

#### Advantages of ggplot2

* consistent underlying grammar of graphics (Wilkinson, 2005)
* plot specification at a high level of abstraction
* very flexible
* theme system for polishing plot appearance
* mature and complete graphics system
* many users, active mailing list

#### What ggplot2 cannot do? 

* 3D visualizations (see the "rgl" package) 
* graph theory type of graphs (node/edges; see "igraph" package)
* interactive graphics (see "ggvis" package)

Website : http://ggplot2.org


```{r,echo=FALSE}
gambling.data <- read.csv(file = "http://data.justice.qld.gov.au/JSD/OLGR/20170817_OLGR_LGA-EGM-data.csv",
                 header = TRUE,
                 sep = ",",
                 stringsAsFactors = FALSE)

class(gambling.data)

names(gambling.data) <- c("Month.Year",
                 "Local.Govt.Area",
                 "Approved.Sites",
                 "Operational.Sites",
                 "Approved.EGMs", 
                 "Operational.EGMs", 
                 "Player.Money.Lost")


```


```{r,echo=FALSE}
#Add a day of month (1st)
date.string<-paste0("1 ",gambling.data$Month.Year)

#Convert to POSIXlt
strptime(date.string, format = "%d %B %Y")->gambling.data$Date

# Add column for year
gambling.data$Year <- as.numeric (format(gambling.data$Date,"%Y"))

# Add column for month
gambling.data$Month <- as.numeric (format(gambling.data$Date,"%m")) 

```

```{r,echo=FALSE}
# Subset LGA
brisbane.only<- gambling.data[gambling.data$Local.Govt.Area=="BRISBANE",]

# Subset by Year & Month
Subset.July2017<-gambling.data[gambling.data$Month.Year=="July 2017",]


```

#### Aggregate gambling.data by Local.Govt.Area & by Year

```{r}

gambling.data <- na.omit(gambling.data)

# Aggregate dataset by taking mean of Player.Money.Lost By each Year

gambling.avg.year <- aggregate(gambling.data$Player.Money.Lost, 
                         by=list(Year = gambling.data$Year), 
                         FUN=mean,
                         na.rm = TRUE)
names(gambling.avg.year)[names(gambling.avg.year) == 'x'] <- 'Avg.Money.Lost'

# Aggregate dataset by taking mean of Player.Money.Lost By Local.Govt.Area

gambling.avg.LGA <- aggregate(gambling.data$Player.Money.Lost, 
                         by=list(Local.Govt.Area = gambling.data$Local.Govt.Area), 
                         FUN=mean,
                         na.rm = TRUE)

names(gambling.avg.LGA)[names(gambling.avg.LGA) == 'x'] <- 'Avg.Money.Lost'

```

### Key components

Every ggplot2 plot has three key components:

1. data,
2. A set of aesthetic mappings between variables in the data and visual
properties, and
3. At least one layer which describes how to render each observation. Layers
are usually created with a geom function.

Let's take a look at a simple example of scatter plot.

### Basic Scatter plot 

```{r}
# Scatter plot of Year and operational sites 
ggplot(gambling.avg.year,
       aes(x=Year,
           y=Avg.Money.Lost
           ))+
geom_point() 

```

This produces a scatterplot defined by:

1. Data: brisbane.only
2. Aesthetic mapping: Date mapped to x position, Operational.Sites to y position (aes() function).
3. Layer: points (geom_point() function)

### Scatter plot with colour added

```{r}
# Scatter plot of Year and operational sites (Brisbane dataset) with Date formatted, x label & y label.

ggplot(gambling.avg.year,
       aes(x=Year,
           y=Avg.Money.Lost
           ))+
geom_point(colour="red") 

```


### Scatter plot with smoothing layer

Now we add a smoothing layer using geom_smooth(method='lm'). Since the method is set as lm (short for linear model), it draws the line of best fit.
The line of best fit is in blue.

```{r}
# Scatter plot of Year and operational sites (with) geom_smooth()
ggplot(gambling.avg.year,
       aes(x=Year,
           y=Avg.Money.Lost
           ))+
geom_point(colour="red") +
geom_smooth(method='lm') 
```

### Barplot

Let's create a barplot with average money lost for each Local.Govt.Area using whole gambling dataset.

```{r}
# Barplot of money lost(average) for each LGA
ggplot(gambling.avg.LGA,
       aes(x=Local.Govt.Area,
           y=Avg.Money.Lost
           ))+
geom_bar(stat="identity") +
theme(axis.text.x = element_text(angle=90,hjust=1)) 

```


### Barplot with ordering

```{r}
ggplot(gambling.avg.LGA,
       aes(x=reorder(Local.Govt.Area,-Avg.Money.Lost),
           y=Avg.Money.Lost
          ))+
geom_bar(stat="identity", fill="blue") + 
theme(axis.text.x = element_text(angle=90,hjust=1)) +
scale_y_continuous(labels = dollar) + 
xlab("Local Government Area") +
ylab("Total Money lost")
```

### Box plot

```{r}
# Box plots for whole dataset by LGA

ggplot((gambling.data), aes(x = reorder(Local.Govt.Area, -Player.Money.Lost, FUN = median), y = Player.Money.Lost)) +
        geom_boxplot() +
        scale_y_continuous(labels = dollar) + 
        theme(axis.text.x = element_text(angle=90,hjust=1)) +
  xlab("Local Government Area") +
  ylab("Total Money lost")

```


Adapted from :

http://ggplot2.org (Book)

http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html




