---
title: "The Brain & Mind Centre"
subtitle: "Day 1"
author: "Sydney Informatics Hub"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: lumen
    hightlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)
```

## Introduction

Welcome to the home page for the Brain & Mind Centre R course. It is a three day course covering six seperate topics that occur frequently to anyone doing pragmatic data analysis. The content is structured in such a way that the material in a section is independent to the other sections, so if you can only make part of the course this won't impact your learning.

These courses are designed to be informal and free flowing, so whilst there is some material we have prepared, if you have a question then at anytime please let us know and we will be more than happy to assist. If there is anything that we can't answer on the day, we will note it down and make sure that we get back to you with a thorough solution.

### Installation & Setup

Below is a script that installs packages that you will need for the next three days and also sets some options which will make the course run smoothly. We will explain what the script does as we get further into the material, but for the meantime we would ask that you copy the script and run it in your environment.

```{r installation, eval=FALSE, message=FALSE}
LoadPackages <- function(x){ for( i in x ){ if( ! require( i , character.only = TRUE ) ){
  install.packages( i , dependencies = TRUE , repos = "http://cran.us.r-project.org")
  require( i , character.only = TRUE )}}}

#  Then try/install packages...
LoadPackages(c("tidyverse",
               "xlsx",
               "ggplot2",
               "foreign",
               "dplyr"))

#options(max.print = 100)

options(scipen = 999)
set.seed(5)
```

## About R

### The R Programming Language

[R](https://www.r-project.org/) is an open source programming language and software environment for statistical computing and graphics that is supported by the R Foundation for Statistical Computing. Today, it is one of the most popular languages, being used all across the world in a wide variety of domains and fields.

### RStudio

[RStudio](https://www.rstudio.com/) is an integrated development environment (IDE) for R. It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, debugging and workspace management.

## Getting Started

### Calculating things in R

Standard math functions work in R:
```{r}
2+3

1e5-5e4

1/1000

sqrt(2)

2*pi
```

Let's use vectors:

```{r}
1:10

2*(1:10)

2^(1:10)
```

We can store values:

```{r}
x<-3
x=3
3->x
x
x^2
```

### Vectors

We can store vectors:

```{r}
x<-1:10
x
2^x

y<-c(3,4,5)
y^2

some.fruits<-c("apple","orange","banana")
some.fruits
```

### Working with Simple Data

A lot of the time in R we are working with tables of data, know as "data frames". 

Commonly, 

**rows** may represent instances e.g. *data points*, *people*, *events*, etc. while 

**columns** will represent different types of data associated with each data point or instance e.g. *Name*, *ID*, *location*, *time*, *value*...

Here is an example data frame

```{r}
simple.data<-data.frame(first.name=c("Alice",
                                     "Bob",
                                     "Cathy",
                                     "Daniel"),
                        gender=as.factor(c("Female",
                                           "Male",
                                           "Female",
                                           "Male")),
                        favourite.number=c(4,23,pi,exp(1)),
                        favourite.letter=c("a","c","x","q"),
                        favourite.weekday=c("Monday","Thursday","Sunday","Friday"),
                        stringsAsFactors = FALSE)
```

### Viewing The Data
Use the function `View()` to visually look at the data

```{r eval=FALSE}
View(simple.data)
```

How many rows do we have?
```{r}
nrow(simple.data)
```
How many columns do we have?
```{r}
ncol(simple.data)
```


### Accessing Subsets
Return only the first 3 rows of the data set
```{r}
simple.data[1:3,]
simple.data[c(1,2,3),]
head(simple.data,3)
```
Return the last two rows in a data set
```{r}
simple.data[nrow(simple.data)+(-1:0),]
tail(simple.data,2)
```
Return a random two rows from the data set
```{r}

dplyr::sample_n(simple.data,2)
```


Return only the "favourite.letter" column in the data set
```{r}
simple.data[,4]
simple.data[,"favourite.letter"]
simple.data[,c("favourite.letter")]
simple.data[,names(simple.data) %in% c("favourite.letter")]
simple.data$favourite.letter
```

Return only the first 3 rows and columns 2 and 5 of the data set
```{r}
simple.data[1:3, c(2,5)]
```

Return the columns named "first.name" and "favourite.number"
```{r}
simple.data[,c("first.name","favourite.number")]
```

### Subsetting the data

## More Realistic Data

The dataset that we will be using today is taken from the Queensland Government website and is [Gaming Machine Data by Local Government Areas](https://data.qld.gov.au/dataset/gaming-machine-data-by-local-government-areas). It has the following variables:

Variable | Definition
-----------------|------------------------------------------------
Month Year | The month and year from which the gambling data is provided.
LGA Region Name | The name of the Local Government Area.
Approved Sites | The number of venues approved to operate electronic gaming machines.
Operational Sites | The number of venues that were operating electronic gaming machines on the last day of the relevant month.
Approved EGMs | The maximum number of electronic gaming machines the venue is approved to operate.
Operational EGMs | The number of electronic gaming machines operating at the venue on the last day of the revelant month.
Metered Win | The amount of money lost by players of eletronic gaming machines.

If you haven't already, the data can be downloaded from the homepage of this course, or alternatively you can click [here](http://data.justice.qld.gov.au/JSD/OLGR/20170817_OLGR_LGA-EGM-data.csv).

### Reading in Data

The first thing to do before we can begin an analysis is loading some data. To do this we can use the below command.

```{r read_data_1}
gambling.data <- read.csv(file = "http://data.justice.qld.gov.au/JSD/OLGR/20170817_OLGR_LGA-EGM-data.csv",
                 header = TRUE,
                 sep = ",",
                 stringsAsFactors = FALSE)
```

Before going any further, let's break down what this command does specifically.

1. It reads the file, which could be 
- a file path e.g. `"~/Documents/MyFolder/datafile.csv"`
- a url like we have here e.g. `"http://www.website.com/subdomain/datafile.csv"`
2. It specifies that the first row in the csv file contains "header" information i.e. column names
3. It specifies that neighbouring columns are separated by a comma ","
4. It says not to convert charachter strings (i.e. text) to something called a factor

How would we find this out if we didn't know already? Look at the help file:
```{r eval=FALSE}
?read.csv
```

Comma Separated Value (.csv) files are usually the standard, simplest format which is compatible with all sort of different software, e.g. R, python, Excel, MATLAB, ...

But if we needed to read in a different format of data, there's usually a function or a library for doing just that,

e.g.
in base R:

- `read.csv()`
- `read.table()`

In the package "xlsx":

- `read.xlsx()`

In the package "foreign":

- `read.dta()` for STATA (version 5-12) data files 
- `read.spss()` for SPSS files

How can we examine this data set now that we've loaded it?



### Viewing The Data
Use the function `View()` to visually look at the data

```{r eval=FALSE}
View(gambling.data)
```

### Changing The Column Names
Using the background information on the dataset, change the column names so they make sense.


```{r changeNames}
class(gambling.data)

names(gambling.data) <- c("Month.Year",
                 "Local.Govt.Area",
                 "Approved.Sites",
                 "Operational.Sites",
                 "Approved.EGMs", 
                 "Operational.EGMs", 
                 "Player.Money.Lost")
```

Afterwards, view the data again to check the column names have been changed.

```{r eval=FALSE}
View(gambling.data)
```



### Accessing Subsets
Return only the first 10 rows of the data set
```{r}
gambling.data[1:10,]
head(gambling.data,10)
```
Return the last ten rows in a data set
```{r}
gambling.data[nrow(gambling.data)+(-9:0),]
tail(gambling.data,10)
```



Return only the period column in the data set (wrapped in head() to avoid a really long output)
```{r}
head(gambling.data[,3],50)
```

Return only the first 5 rows and columns 4 through 7 of the data set
```{r}
gambling.data[1:5, 4:7]
```

Return the column named "Player.Money.Lost"
```{r}
head(gambling.data[,"Player.Money.Lost"],50) 
```

Return the column named "Player.Money.Lost"
```{r}
head(gambling.data$Player.Money.Lost, n=50)
```

Return the columns named "Local.Govt.Area" and "Player.Money.Lost"
```{r}
head(gambling.data[,c("Local.Govt.Area","Player.Money.Lost")],30)
```

### Summarising The Data
Use the summary function to return a quick summary

```{r}
summary(gambling.data)
```

Use the summary function to return a quick summary of only the third column

```{r}
summary(gambling.data[,3])
```

### Data Types
Check the data class of the third column
Convert the third column to factor using the factor() function
Use the summary function to return a quick summary of only the third column, what changed?

```{r}
class(gambling.data[,3])

gambling.data[,3] <- factor(gambling.data[,3])

summary(gambling.data[,3])

```

### Creating Vectors
Create a vector called "vector" with the information from the first record

```{r}
vector <- gambling.data[1,]
```

### Adding records
Add vector to the data set using the rbind() function
Add a new column called "source" to the data set, where each record is "Wankarani Settlement Dataset"
Add a new column called "author" to the data set, where each record is "Timothy L. McAndrews"

```{r}
gambling.data[136,] <- vector
gambling.data$source <- "Wankarani Settlement dataset"
gambling.data$author <- "Timothy L. McAndrews"
```


### Writing Data
Use the write.table() function to make the datafile a text file

```{r}
write.table(x = gambling.data,
            file = "gamblingData.txt",
            row.names = FALSE,
            col.names = TRUE,
            sep = ",")
```

Use the write.csv() function to make the datafile a csv (comma separated values) file

```{r}
write.csv(x = gambling.data,
          file = "gamblingData.csv",
          row.names = FALSE,
          col.names = TRUE,
          sep = ",")
```

          
Use the write.xlsx