---
title: "The Brain & Mind Centre"
author: "Sydney Informatics Hub"
subtitle: Day 2
output:
  html_document:
    hightlight: github
    theme: lumen
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
---

<!-- https://stackoverflow.com/questions/46201753/rmarkdown-indentation-of-toc-items-in-html-output -->
<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)
```

# Controls & Loops

These allow you to control the flow of execution of a script typically inside of a function. Common ones include:

* if, else
* for
* while
* repeat
* break
* next
* function

## Control statements

### If statement
```{r, eval=FALSE, message=FALSE}

if (condition)
{
  # do something
} 
```

Sometimes, we want to do something even if the if-statement returns FALSE. In this case, we use an else statement.

### If else statement
```{r, eval=FALSE, message=FALSE}

if (condition)
{
  # do something
} 
else 
{
  # do something else
}

```

Hadley Wickham has published a style guide for R http://adv-r.had.co.nz/Style.html. 

"An opening curly brace should never go on its own line and should always be followed by a new line. A closing curly brace should always go on its own line, unless it’s followed by else. Always indent the code inside curly braces.""

#### If else statements with better styling
```{r, eval=FALSE, message=FALSE}

if (condition) {
  # do something
} else {
  # do something else
}

```

An example of if..else statement:

```{r}
x <- 1
if (x > 1) {
  print("x is greater than 1")
} else {
  print("x is less than or equal to 1")
} 
```


### If, else if and else statements
```{r, eval=FALSE, message=FALSE}

if (condition1) {
  # do something
} else if (condition2) {
  # do something else
} else {
  # do something different
}

```

For example:

```{r}
x <- 1
if (x > 1) {
  print("x is greater than 1")
} else if ( x < 1) {
  print ("x is less than 1")
} else {
  print("x is equal to 1")
} 
```


#### Quiz
Write a set of if..else statements that will test a variable "mhi5_score". Print out "depression" if it is less than 52 and "control" if it is greater than or equal to 52
```{r}
mhi5_score <- 45
# put your if else statements here

```


## Loops

### For loops

A for-loop works on an iterable variable and assigns successive values till the end of a sequence.

```{r}
for (i in 1:10) {
  print(i)
}

```

```{r}
x <- c("apples", "oranges", "bananas", "strawberries")

for (i in 1:4) {
  print(x[i])
}

```

For short-hand coding, we can put the contents of the loop on the same line as the for-statement

```{r}
for (i in 1:4) print(x[i])
``` 

We can also simply print out each element of our vector using a for-loop to traverse it

```{r}
for (i in x) print(i)
```


### For loop examples

We've seen that we can use for loops for vectors. We can also use for loops for data frames

```{r}
# read in the gambling data from Day 1
gambling.data <- read.csv(file = "http://data.justice.qld.gov.au/JSD/OLGR/20170817_OLGR_LGA-EGM-data.csv",
                 header = TRUE,
                 sep = ",",
                 stringsAsFactors = FALSE)

# rename columns
names(gambling.data)[2] <- "Local.Govt.Area"
names(gambling.data)[7] <- "Player.Money.Lost"

#Add a day of month (1st) to each date string
date.string <- paste0( "1 " , gambling.data$Month.Year )

#Convert to POSIXlt, a date-time format
strptime( date.string , format = "%d %B %Y" ) -> gambling.data$Date

# subset to Brisbane only 
brisbane.only <- gambling.data[gambling.data$Local.Govt.Area=="BRISBANE",]
row.indicies <- (brisbane.only$Date>="2010-01-01 AEST" &
                 brisbane.only$Date<="2010-12-31 AEST")

(brisbane.2010.data <- brisbane.only[row.indicies,])
```

We can traverse across each row of the data frame

```{r}
# for each row in brisbane.2010.data, print the player money lost
numrows <- nrow(brisbane.2010.data)
for (i in 1:numrows) {
  print(brisbane.2010.data[i, "Player.Money.Lost"])
}
```


Or across each column of the data frame

```{r, message=FALSE}
# for each column in brisbane.2010.data, print the data type using the class() function
numcols <- ncol(brisbane.2010.data)
for (i in 1:numcols) {
  columndata <- brisbane.2010.data[, i]
  print(class(columndata))
}
```


#### For loops and if statements
We can incorporate our knowledge of if..else statements with for-loops. 

```{r}
# for each row, print those where the Player.Money.Lost is greater than 32 million
numrows <- nrow(brisbane.2010.data)
for (i in 1:numrows) {
  if (brisbane.2010.data[i, "Player.Money.Lost"] > 32000000){
    print(brisbane.2010.data[i, "Player.Money.Lost"])
  }
}
```


#### Quiz
For each row of the brisbane.2010.data dataframe, write code to test whether the Player.Money.Lost value is greater than 32 million. If so, print "Greater than 32 million". Otherwise, print "Not greater than 32 million"

```{r}
numrows <- nrow(brisbane.2010.data)
for (i in 1:numrows) {
  # write your code here 
}
```




### While loop
```{r}
i <- 1
while (i < 10) {
  print(i)
  i <- i + 1
}
```

Make sure there is a way to exit out of a while loop. Otherwise, you can get stuck in an infinite loop. Below is an example of one.

```{r, eval=FALSE, message=FALSE}
i <- 1
while (i < 10) {
  print(i)
}
```

### Break statement

A break statement is used in a loop to stop the iterations and flow the control outside of the loop.

```{r}
x = 1:10 
for (i in x){ 
  if (i == 2){ 
    break 
  }
  print(i)
}
```

### Next statement

Next statement enables to skip the current iteration of a loop without terminating it.

```{r}
x = 1:4 
for (i in x) { 
  if (i == 2){ 
    next
  }
  print(i)
}
```

## User-defined functions

If you have to repeat the same few lines of code more than once, then you really need to write a function. Functions are a fundamental building block of R. You use them all the time in R and it's not that much harder to string functions together (or write entirely new ones from scratch) to do more.

* R functions are objects just like anything else.
* By default, R function arguments are lazy - they're only evaluated if they're actually used:
* Every call on a R object is almost always a function call.

#### Basic components of a function

* The body(), the code inside the function.
* The formals(), the "formal" argument list, which controls how you can call the function.
* The environment() which determines how variables referred to inside the function are found.
* args() to list arguments.

```{r}
myFunction <- function(parameter1, parameter2) {
  # do something with parameter1 and parameter2
  value <- parameter1 + parameter2
  return(value)
}
```

Functions should have good descriptive names

```{r}
# Create a function to perform addition
add <- function(num1, num2) {
  return(num1 + num2)
}

# Call the add function
add(2,5)
```


#### Example 1 : Loop through brisbane.2010.data and sum up the Player.Money.Lost

```{r}
total.money.lost <- 0

for (i in 1:nrow(brisbane.2010.data)) {
  total.money.lost = total.money.lost + brisbane.2010.data[i, "Player.Money.Lost"]
}

print(total.money.lost)
```


#### Example 2 : Create a function to sum up money lost by passing vector of player.money.lost values

```{r}
sumMoneyLost <- function(some.gambling.dataframe){
  
  total.money.lost = 0

  for (i in 1:nrow(some.gambling.dataframe)) {
    total.money.lost = total.money.lost + some.gambling.dataframe[i, "Player.Money.Lost"]
  }
  return(total.money.lost)
}

# Call function sum.money.lost
sumMoneyLost(brisbane.2010.data)

```

With user-defined functions, you won't need to repeat your code over and over again. You can just call it for another dataset

```{r}
sumMoneyLost(brisbane.only)
```


If, however, there is already a function to perform what you need, save time and use that instead.

```{r}
# This will return NA
sumMoneyLost(gambling.data)

# We will use the sum() function instead which has a na.rm option
sum(gambling.data$Player.Money.Lost, na.rm = TRUE)
```






Adapted from :

https://ramnathv.github.io/pycon2014-r/learn/controls.html

https://www.r-bloggers.com/control-structures-loops-in-r/
